Can historical data predict population responses to climate change experiments?

### 2016-12-19 precip/manuscript/precip_ms.tex
adding more manuscript revisions.  Struggled to get the soil moisture treatment effects script to work.  Need more work on that

@@

As in the survival regression above, specific specific parameters describing the treatment effects on growth are added in the treatment model, 
\begin{equation}
\boldsymbol{T} \chi_{j}^G 
\label{eqn:growT}
\end{equation}

where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.

@@

We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}. The prior distributions and model code used are described more completely in appendix A. Each model was run for 2,000 iterations and four independent chains with different initial values for parameters. We discarded the initial 1,000 samples. Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan manual}.

@@

The vital rate regressions allow us to evaluate whether soil moisture and the experimental treatments had an effect on species performance.  But the population response ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.  To evaluate whether the climate models could predict the overall population response of these species in the drought and irrigation experiment we used an individual-based model (IBM) to compare observed and predicted changes in population size, defined as plot-level percent cover, from one year to the next in the experimental plots.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the observed genet sizes and locations of the four focal species observed in year $t$ in each quadrat. For every individual genet in a quadrat, we projected its size and survival probability in the next year using the growth and survival models and the appropriate crowding and soil moisture or treatment covariates for that year and quadrat.  Likewise we projected the number of new recruits in the quadrat in the next year using the recruitment model. We calculated the expected cover in year $t+1$ as the sum area of new recruits, plus the sum of the expected area of existing plants multiplied by each plant's expected survival probability.

We accounted for the uncertainty in our random year effects in our projections, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}. We generated predictions using the full posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions. Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the models (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a negative binomial distribution with a mean of $\lambda$).

@@

We successfully changed the average soil moisture measured in the plots we covered with drought shelters and irrigated, in the directions we expected (fig); our spot measurements conducted during the springs of X, X and X showed that drought plots were X percent drier, while irrigation plots were X percent wetter than ambient plots.  This difference was highly significant () but there was a lot of spatial variation in the soil moisture measurements within treatments (\ref{fig:}). The automatically recorded soil moisture data from only half of the experimental plots also showed that the drought treatment decreased and the irrigation treatment increased soil moisture, but the size of these effects were dependent on season and whether it had rained recently (). The best model for predicting the average treatment effects on soil moisture was : .  We did not find evidence that the drought shelters and the irrigation treatments significantly changed air temperature within the plots ().

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=1\textwidth]{treatment_trends_cover}
\caption{Cover trends by treatment for the four dominant species.  Note the difference in y-axis scale between the canopy cover of \textit{A. tripartita} and the basal cover of the dominant grasses. ``Shrub removal'' refers to removal of \textit{A. tripartita}, and ``Grass removal'' refers to removal of all perennial grasses. The cover of species in treatments from which they are removed (e.g. \textit{A. tripartita} cover in shrub removal plots) is included to show values at the start of the experiment in 2011 and the efficacy of the removal treatments.}
\label{fig:CoverTrends}
\end{figure}

@@

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{treatment_tests}
 \caption{Effects of removal treatments on the survival, growth, and recruitment of the four dominant species. Circles show the estimated coefficient of the removal effect, bars show 95\% confidence intervals. The removal effect shows response of \textit{A. tripartita} ("ARTR") to perennial grass removal, and response of the perennial grasses to \textit{A. tripartita} removal, after accounting for local neighborhood interactions. }
 \label{fig:VitalRateTest}
 \end{figure}

@@

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step}
 \caption{Observed and predicted cover for each of the four modeled species. Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats. Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients. }
 \label{fig:IBM1step}
 \end{figure}

@@

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{boxplots}
 \caption{Equilibrium cover of the four dominant species simulated by the IPM. Boxplots show interannual variation reflecting random year effects. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), and green boxes show results from a model with  \textit{A. tripartita} set to zero and removal treatment coefficients included.  }
 \label{fig:IPMresults}
 \end{figure}

@@

@@

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{CompKernels}
 \caption{Competition kernels for the four dominant species.  }
 \label{fig:CompKernels}
 \end{figure}

  \begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{climate}
  \caption{Annual precipitation (a) and mean temperature (b) during the period of the experiment, shown against the long-term (1927-2016) means (solid blue and red lines) and 5\% and 95\% quantiles (dashed lines).  }
  \label{fig:climate}
  \end{figure}

  \begin{figure}[tbp]
  \centering
  \includegraphics[width=1\textwidth]{PSSP_W_scatters}
  \caption{Bivariate scatter plots comparing crowding exerted by each neighbor species on \textit{P. spicata}. Blue symbols show values from controls plots, red symbols show values from \textit{Artemisia} removal plots.``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.  }
  \label{fig:Wscatters}
  \end{figure}

  \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{PSSP_W_byTrt}
 \caption{Distribution of crowding exerted by each neighbor species on \textit{P. spicata} individuals in controls plots (blue) and removal quadrats (red) in 2011, before \textit{Artemisia}  removals were conducted. Inset bar graphs show the probability that a particular neighbor species was present ($W>0$) in a focal plant's local neighborhood in control (blue) and removal (red) quadrats. ``W.allcov'' refers to the aggregated crowding by all shrubs and perennial grasses beyond the focal dominant species, and ``W.allpts'' refers to the aggregated crowding by forb species.  }
 \label{fig:W-by-treatment}
 \end{figure}

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}
 \caption{Observed and predicted cover for each of the four modeled species assuming maximum effects of removal. Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats. Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients representing the outermost values of the 95\% confidence intervals. }
 \label{fig:IBM1step-maxCI}
 \end{figure}

  \begin{figure}[tbp]
  \centering
  \includegraphics[width=1\textwidth]{boxplots-maxCI}
  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects. Boxplots show interannual variation reflecting random year effects. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), green boxes show results from a model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.  }
  \label{fig:IPMresults-maxCI}
  \end{figure}

>>>>

\graphicspath{{../figures/}}

@@

As in the survival regression above, specific specific parameters describing the treatment effects on growth are added in the treatment model, 
\begin{equation}
\boldsymbol{T} \chi_{j}^G 
\label{eqn:growT}
\end{equation}
where $\chi$ is a treatment effect describing the effect of experimental treatment $h$ on growth, including treatment by size interactions.

@@

We fit all vital rate models using Hamiltonian-Markov Chain Monte Carlo (HMCMC) simulations in STAN 6.4 \citep{stan} and  \citep{rstan}. The priors and model code are described more completely in appendix A. Each model was run for 2,000 iterations and four independent chains with different initial values for parameters. We discarded the initial 1,000 samples. Convergence was observed graphically for all parameters, and confirmed by assessing the split $\widehat{R}$ statistic which at convergence is equal to one \citep{stan manual}.

@@

The vital rate regressions allow us to evaluate whether soil moisture and the experimental treatments had an effect on species performance.  But the population response ultimately depends on the integrated effects of treatment or soil moisture on all three vital rates.  To evaluate whether the climate models could predict the responses of these species in the drought and irrigation experiment at the overall population level we used an individual-based model (IBM) to compare observed and predicted changes in population size from one year to the next.

To simulate changes in cover in each quadrat from year $t$ to year $t+1$, we initialized the IBM with the observed genet sizes and locations of the four focal species observed in year $t$ in each quadrat. For every individual genet in a quadrat, we projected its size and survival probability in the next year using the growth and survival models and the appropriate crowding and soil moisture or treatment covariates for that year and quadrat.  Likewise we projected the number of new recruits in the quadrat in the next year using the recruitment model. We calculated the expected cover in year $t+1$ as the total area of new recruits, plus the sum of the predicted area of each existing plant at time $t+1$ multiplied by each plant's expected survival probability from time $t$ to $t+1$.

We accounted for the uncertainty in our random year effects when generating predictions, by drawing random year effects for each predicted year from a normal distribution with a mean of zero and a standard deviation drawn from the posterior estimate of the standard deviation of the random year effects \ref{eqn:}. We generated predictions using the full posterior distributions of each model parameter which allowed us to carry forward all of the uncertainty of the fitted vital rate models into our cover predictions. Because we were interested in comparing model predictions to observations, and were not interested in the effects of demographic stochasticity, we used a deterministic version of the models (e.g., recruitment is the $\lambda$ of (\ref{eqn:recrProcessModel}), rather than a random draw from a negative binomial distribution with a mean of $\lambda$).

@@

Our treatments successfully changed the soil moisture in the experimental plots in the directions expected (fig. \ref{fig:VWCspot}). Our spot measurements of soil moisture from the spring showed that drought plots were roughly 50\% drier, while irrigated plots were roughly 40\% wetter than ambient plots. This difference was highly significant (table \ref{table:spotVWC}).

The automatically recorded soil moisture data from only half of the experimental plots also showed that the drought treatment decreased and the irrigation treatment increased soil moisture, but the size of these effects were dependent on season and whether it had rained recently (). The best model for predicting the average treatment effects on soil moisture was : .  We did not find evidence that the drought shelters and the irrigation treatments significantly changed air temperature within the plots ().

@@

@@

@@

@@

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% TABLES !
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\begin{table}[h]
	\caption{Spring soil moisture}
	\centering
	%\begin{center}
	\begin{tabular}{l c c }
		\hline
		& Model 1 \\
		\hline
		(Intercept)                  & $8.81^{*}$    	& $[5.78;\ 11.83]$  \\
		TreatmentDrought             & $-3.97^{*}$      & $[-4.84;\ -3.09]$ \\
		TreatmentIrrigation          & $3.26^{*}$       & $[2.39;\ 4.14]$   \\
		\hline
		AIC                          & 3191.87           \\
		BIC                          & 3222.92           \\
		Log Likelihood               & -1588.93          \\
		Num. obs.                    & 624               \\
		Num. groups: plot            & 24                \\
		Num. groups: PrecipGroup     & 8                 \\
		Num. groups: date            & 5                 \\
		Var: plot (Intercept)        & 0.45              \\
		Var: PrecipGroup (Intercept) & 0.23              \\
		Var: date (Intercept)        & 11.24             \\
		Var: Residual                & 8.90              \\
		\hline
		\multicolumn{2}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
	\end{tabular}
	\label{table:spotVWC}
	%\end{center}
\end{table}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% FIGURES !
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{VWC_spot_measurements}
	\caption{Soil moisture in the upper 5 cm of drought and irrigated plots compared to ambient controls. Soil moisture was measured at six locations around each plot at five different dates during the spring. Control plots were nearby areas of experiencing ambient soil moisture. Box plots show the mean soil moisture and the upper lower interquartile range.  Dots show individual soil moisture measurements. Readings of volumetric soil moisture less than zero were occasionally obtained in very dry soil.}
	\label{fig:VWCspot}
\end{figure}

\begin{figure}[tbp]
	\centering
	\includegraphics[width=1\textwidth]{avg_daily_soil_moisture}
	\caption{Average soil moisture in the control, drought, and irrigation treatments during each year of the experiment.  Soil moisture was monitored in four drought plots and four irrigated plots and four ambient control plots.  Averages represent average soil moisture recorded every two hours over the course of the day at 5 cm and 25 cm depths.}
	\label{fig:CoverTrends}
\end{figure}

@@
